A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-12 PPSN2016/paper.Rnw
Reinserting the table

\begin{table}[htb]
\caption{Experiment table, with summary of results. \label{tab:runs}}
\begin{center}

<<experiments, cache=FALSE,echo=FALSE>>=
experiment.aggregates$Experiment <- c("base","graph1","graph2","graph3","reboot","cachereboot")
kable(experiment.aggregates)
@ %def

@@

The table shows that every run included more than 30 experiments. The
%Pablo: reference to table? (maybe the final print will misplace it)
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment, with a
number around 50 in the second, and most recent, batch of experiments. 
%
\begin{table}[htb]
\caption{Summary of time per run, number of IPs and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|cccc}
\hline
     & \multicolumn{2}{c}{IPs} & \multicolumn{2}{c}{Median} \\
Experiment & Median & Max & time (s) &  \#{\tt  PUT}s \\
\hline
4/4 & 5 & 16 & 2040 & 18   \\
4/24 &  5 & 29 & 732 & 11  \\
7/31 & 5 & 14 & 260 & 23   \\
\hline
Cache=128 & 5 & 17 & 222.2 & 124 \\ 
Cache=64 & 8 & 38 & 51.3 & 100 \\
Cache=32 & 6 & 19 & 58.9 & 45 \\
\hline
% Data in .RData file in Splash-volunteer
\end{tabular}
\end{center}
\end{table}
%

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of IPs
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP. The
first striking result is that in all cases, 50\% of the 
experiments involved 5 or less IPs. This is consistent with previous results
\cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the expected number of volunteer IPs. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work and does not vary across the two different batches.

We will have to analyze differently the median time, since the two
%Pablo: different verbal times ("will have", and before, "is shown"). Homogenize.
batches are solving different problems. In both cases it possesses a
big range of variation, but 50\% of 
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. Remarkably
enough, the time is more consistent in the second batch and always
around one minute, in two cases even less, and that happens when the
median number of IPs is higher. It should be noted that while the
first batch of experiments took several days in each case, the second
only lasted for a few hours, with a more continued effort of
publicizing it in social networks. This is specially true in the
%Pablo: removed semicolon before "This". @unintendedbear does not like them :P
case of cache equal to 64, which is noticed by the high number of
volunteers participating in the experiment. The conclusion is that,
in general, the key factor in the time needed to find the solution
is, as expected, the number of volunteers it is able to gather on a
short notice.

The number of {\tt PUTs}, every one corresponding to 100 generations,
is the algorithmic result. It is relatively unchanged for the first
batch and around 20, that is, 2000 generations or 2000*128 = 256000
evaluations. In this case, an ``unlimited'' cache was used, with all
individuals sent from clients stored until the end of the
experiment. However, we were interested in measuring also the
performance of the algorithm itself by changing the cache size, after
making it limited. As it can be seen in the table, there is a clear
%Pablo: which table?
change in the number of evaluations needed, with smaller cache sizes
producing solutions in less evaluations, until it is for the cache
size = 32 roughly twice as much as with the previous problem, with 40
traps. This is a good result and is also algorithmically consistent
with other results obtained using the same type of problems. Since in
this paper we were interested in leveraging the user's CPU cycles by
improving the algorithm, a good conclusion of this paper is that
having a small pool size helps clients to obtain ``good'' individuals
from the pool, as opposed to any individual that could be obtained
before. Besides, cache policy deletes the oldest individuals, which
makes those in the pool be {\em current}, helping then newcomers and %Pablo: "then helping newcomers", instead?
any participant obtain the best individuals found in the last part of
the experiment. Besides, a limited cache helps also in cases with a
bigger search space or longer running times when the server simply
crashed due to lack of RAM.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-alife-128.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-alife-64.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-alife-32.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of blue indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right and top to bottom experiments 4/4, 4/24 and 7/31,
followed by experiments with 50 traps, cache=128, 64, 32.}
\label{fig:duration}
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
the probability of someone new joining them is higher. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. 
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.95\linewidth]{ips-per-minute-cache=64.png}
\caption{Simultaneous IPs every minute of the experiment with cache =
  64. \label{fig:otisdriftwood}}
\end{figure}

That is why we used a more difficult problem in the second batch of
experiments, which is shown in the bottom row of Figure
\ref{fig:duration}. The pattern is remarkably similar, showing a
positive correlation between the time for solving the problem and the
number of computers, at least for cache sizes 128 and 64. However, it
is interesting to observe that, for cache=32, the time needed to find
the solution decreases from one to approximately 4-5 nodes, to then
increase for a higher number of participating computers, distinguished
by IP. The green dot at the bottom is probably an outlier that we
will try to explain later on. This leads us to conclude that a better
amount of computers might contribute to speed up the solution, if the
time the experiment ideally takes is sufficient, that is, of the order of a
minute, and enough volunteers concur simultaneously. This is also
observed, not so clearly, in the case of cache=64, with an interval of
around 10 IPs obtaining less time than experiments with less or more
IPs, and of the same order, between 10 and 100 seconds. If we look at
the graph that shows the number of IPs or volunteers per minute for
this experiment, shown in Figure \ref{fig:otisdriftwood}, we see that
there are peaks of more than 25 volunteers, and a period of several
hours with a minimum of 6 computers and peaks of more than 10. The
long period after midnight where there is a single volunteer left
masks the success achieved during this set of experiments, from which
we draw two lessons: first, you need a social network influencer to
%Pablo: people know the meaning of "influencer"? Probably yes, but I mention it just in case.
announce your experiments and second, no matter what, do not do any
experiment after midnight. This statement, which might seem tongue in
cheek, in fact, it is a conclusion drawn from the experimental data and
to what extent the social network is an essential part of the
description and performance of the NodIO volunteer computing system.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-alife-128.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-alife-64.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-alife-32.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right and top to bottom, experiments
  4/4, 4/24 and 7/31 and caches=128,64, 32} 
\label{fig:zipf:os}
% First ones made with time-vs-IPs-openshift.R:  
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows an S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds; this S-shape appears too in the experiments with cache=128
(bottom row, left). The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, on the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. However, since they have a greenish color, implying a low
number of IPs, they might be due to clients {\em carrying over} from
the previous one. This is a characteristic of this implementation
which will be examined later on, but at any rate, if we discard those
experiments that take too much or too little, there is a decreasing
exponential distribution that corresponds to the Zipf's law.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{weibull-puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{weibull-puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{weibull-puts-openshift-7-31.png}
\includegraphics[width=0.32\linewidth]{weibull-fit-cache=128.png}
\includegraphics[width=0.32\linewidth]{weibull-fit-cache=64.png}
\includegraphics[width=0.32\linewidth]{weibull-fit-cache=32.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Weibull
  distribution (in red). From left to right and top to bottom, experiments
  4/4, 4/24 and 7/31 and new experiments with cache=128, 64, 32.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
A similar exponential distribution also appears if we rank HTTP {\tt
  PUT}s, equivalents to the number of 
generations divided by 100, or to evaluations divided by 12800,
contributed by every user, which is shown 
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
that is, a power law with a small {\em bump} in the lowest
values. After testing the Generalized Extreme Value distribution and
failing for the new batch of experiments, we have fitted it to a
Weibull distribution \cite{thoman1969inferences} with the resulting
parameters shown in Table \ref{tab:puts:os}. 
%
\begin{table}
\caption{Weibull distribution  parameters of the fit of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|cc}
\hline
Experiment  &  Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  43.07 $\pm$ 5.80 &  0.57 $\pm$ 0.03 \\
4/24 & 22.97 $\pm$ 1.57 & 0.66 $\pm$  0.02  \\
7/31 &  53.18 $\pm$ 7.77 &  0.54 $\pm$ 0.03   \\
\hline
Cache 128 & 205.28 $\pm$ 32.10 & 0.77 $\pm$ 0.07 \\ 
Cache 64 & 178.99 $\pm$ 36.44 & 0.60 $\pm$ 0.05 \\ 
Cache 32 & 168.15 $\pm$ 49.28 & 0.57 $\pm$ 0.07 \\
\hline
% from plot-zipf-openshift.R and ips-puts-alife.R
\end{tabular}
\end{center}
\end{table}
%
The inverse
Weibull distribution is a special case of the GEV distribution we have
used in those papers, and appears usually in natural sciences and
artificial life, usually related to decay. It has been frequently
fitted to volunteer computing frameworks such as SETI@home
\cite{javadi2009mining}. The model that user behavior follows can be
explained straighforwardly: when users visit the page, it draws their
attention for a limited amount of time. They give it a chance for a
few seconds. If something there amuses them or they can engage in a
conversation about it, they stay for a while longer, otherwise, they
leave. The {\em scale} parameter, which is around 20-40 in the first
batch for the 40 Trap problem and between 160 and 200 in the 60 traps
problem, depends mainly on the maximum number of generations people
leave it running. Since it finishes or stalls after a number of
generations shown in Figure \cite{tab:summary:os}, volunteers just
leave after that. Curiously enough, the scale parameter is roughly
twice the median number of {\tt PUT}s per experiment, showing that, on
average, the most loyal users reload the page twice after finishing or
after seeing the evolution does not progress. This rule of thumb
breaks down with the last experiment, however, which is interesting by
itself, too.

The slope or shape parameter, on the other hand, indicates the overall
shape of the curve. A value less than 1 indicates a concave (in a
non-algorithmic scale) curve,
with figures closer to one indicating a smaller slope. In all cases
values are between 0.54 and 0.77, independently of the experiment. It
might be the case that this number depends more on the total number of
experiments carried out, with sets with more experiments, both in the
middle, having values between 0.60 and 0.70. The distribution is
remarkably similar which gives us a model of user behavior that is, to
a certain extent, independent of the experiment.

These experiments show that, as it was proved for other volunteer
computer frameworks and also in the case of games, user engagement
follows a Weibull distribution. This makes engagement the key for
leveraging the performance of the sociotechnical metacomputer and a
way to improve results in the future.

>>>>

\begin{table}[htb]
\caption{Experiment table, with summary of results. \label{tab:runs}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Experiment & Time (ms) & \multicolumn{2}{|c|}{IPs} & \multicolumn{2}{|c|}{PUTs} & \#Runs\\
           &           & Total & Contributing & Total & Contributing & \\
\hline
base & 120056.0 & 10 & 4 & 256.0 & 21.0 & 81\\
\hline
graph1 & 152619.5 & 9 & 4 & 212.5 & 26.5 & 36\\
graph2 & 152619.5 & 9 & 4 & 212.5 & 26.5 & 36\\
graph3 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 36\\
\hline
reboot & 513354.5 & 9 & 6 & 545.5 & 62.5 & 40\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 55\\
\hline
\end{tabular}
%% Initially generated with this code, 
%% <<experiments, cache=FALSE,echo=FALSE>>=
%% experiment.aggregates$Experiment <- c("base","graph1","graph2","graph3","reboot","cachereboot")
%% kable(experiment.aggregates)
%% @ %def

@@
